Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography

### 2024-03-16 manuscript/manuscript.tex
Final revisions before submission

\abstract{It is widely acknowledged that opaque orthographies place additional demands on learning, often requiring many years to fully acquire. It is less widely recognized, however, that such opacity may offer certain benefits in the context of reading. For example, heterographic homophones such as <knight> and <night> (words that sound the same but which are spelled differently) impose additional costs in learning but reduce ambiguity in reading. Here, we consider the possibility that---left to evolve freely---writing systems will sometimes choose to forego some simplicity for the sake of informativeness when there is functional pressure to do so. We investigate this hypothesis by simulating the evolution of orthography as it is transmitted from one generation to the next, both with and without a communicative pressure for ambiguity avoidance. In addition, we consider two mechanisms by which informative heterography might be selected for: \textit{differentiation}, in which new spellings are created to differentiate meaning (e.g., <lite> and <light>), and \textit{conservation}, in which heterography arises as a byproduct of sound change (e.g., <meat> and <meet>). Under pressure from learning alone, orthographic systems become transparent, but when combined with communicative pressure, they tend to favor some additional informativeness. Nevertheless, our findings also suggest that, in the long term, simpler, transparent spellings will tend to be preferred.}

@@

\noindent
Writing systems, particularly those employing alphabetic scripts, are commonly regarded as visual representations of speech, with each letter corresponding to a distinct sound. However, it is also well understood that writing systems diverge from their spoken counterparts in important ways \parencite{Bolinger:1946, Coulmas:1991}. The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech \parencite{Parkes:1992, Saenger:1997}, and it seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality \parencite{Rastle:2019}. In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality. In principle, the same may be true of spelling: Words may be spelled in ways that diverge from the spoken language because such divergence confers some benefit in reading \parencite{Ulicheva:2020}.

One potential case of such functional divergence is heterographic homophony---words that sound alike but which are written differently (e.g., \textit{meat} and \textit{meet}). Heterographic spellings such as these may serve a valuable function in reading. For example, an English speaker faced with a spoken sentence beginning /ðɛr.../ will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr~kat/, a form of the verb \textit{to~be}, as in /ðɛr~ɪz/, or the progressive form of a verb, as in /ðɛr~gəʊɪŋ/. In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on processing the upcoming syntactic structure and semantic content. Heterographic homophony can also occur in affixes. The English suffixes \textit{-er} (denoting the comparative form of an adjective; e.g., \textit{nicer}) and \textit{-or} (denoting the performer of an action; e.g., \textit{actor}) are homophonous in speech (/ər/), but their spellings differentiate these meanings in writing. Of course, English orthography is suboptimal here in~that \textit{-er} may also indicate agentive status (e.g., \textit{builder}); nevertheless, statistical patterns such as these hold across a variety of English affixes \parencite{Berg:2017} and it has been shown that readers are sensitive to and make use of such cues in reading \parencite{Ulicheva:2020}. \newmaterial{These features might be especially important given the differing constraints of the written modality, including the lack of other cues to meaning, such as stress, context, and body language, and the inability for reader and writer to engage in immediate feedback and repair. Furthermore, written language tends to be more sophisticated than spoken language (e.g., longer, more elaborate sentences), placing different pressures on ambiguity resolution.}

\newmaterial{Although heterography is particularly notable in English, it is also a feature of many other languages and writing systems. French is similar to English in having a large number of heterographic homophones; for example, the heterographic words \textit{cent} (\textit{hundred}), \textit{sang} (\textit{blood}), \textit{sans} (\textit{without}), and \textit{sens} (\textit{feel}) are all pronounced /sɑ̃/ (although their pronunciations will sometimes be distinguished through liaison). In Danish, the words \textit{hver} (\textit{every}), \textit{vejr} (\textit{weather}), \textit{vær} (\textit{be}), and \textit{værd} (\textit{worth}) are all pronounced /vεˀɐ̯/. In Vietnamese, the graphemes <d>, <gi>, and <r> are homophonous, resulting in sets like \textit{dao} (\textit{knife}), \textit{giao} (\textit{delivery}), and \textit{rao} (\textit{advertise}), all pronounced /zāw/. In some cases, features of an orthography designed for other purposes can inadvertently result in homophone disambiguation: Noun capitalization in German contrasts \textit{Wagen} (\textit{car}) and \textit{wagen} (\textit{to dare}), both pronounced /vaːɡṇ/; eclipsis marking in Irish contrasts \textit{bpáistí} (\textit{children}) and \textit{báistí} (\textit{rain}), both pronounced /bɑʃti/; and the morphological principle of Russian orthography contrasts \textit{приступить} (\textit{to start}) and \textit{преступить} (\textit{to transgress}), both pronounced /pristupʲitʲ/. Even in the most transparent of orthographies it is possible to find some instances of heterography: In Italian, the residual~\textsc{h} inherited from Latin contrasts \textit{hanno} (\textit{have}) and \textit{anno} (\textit{year}), both pronounced /anno/, while the grave accent is sometimes used to distinguish common homophonous words, such as \textit{la} (\textit{the}) and \textit{là} (\textit{there}).}

\newmaterial{However, perhaps the most elaborate example of how a writing system can deal with homophony head-on is the Chinese script. The Chinese spoken languages are rich in homophones, making heterographic spellings---and therefore a logographic script---particularly useful \parencite{Katz:1992}. In Mandarin Chinese, the words 糖 (\textit{sugar}), 塘 (\textit{embankment}), 溏 (\textit{pond}), and 搪 (\textit{to block}) are homophonous in speech but heterographic in writing---the phonetic radical on the right (唐; /táŋ/) represents the spoken syllable, while the semantic radicals on the left differentiate the meanings. In addition, 唐 itself is a surname/dynasty (\textit{Tang}), and another unrelated word 堂 (\textit{hall}) is also pronounced /táŋ/, yielding at~least six ways to write the same sound depending on the meaning. This allows written Chinese to convey more information about meaning---to be more \textit{informative}---than its spoken counterpart.}

Despite the benefits that heterography may provide in reading, it comes with two main costs. Firstly, by definition, heterography implies that a single sound can be spelled multiple ways. In English, the heterographic spellings <meat> and <meet> imply that /i/ can be spelled <ea> or <ee>. Readers are therefore required to learn alternate spellings for a single sound, \newmaterial{resulting in longer learning periods and more difficult decoding} \parencite{Reis:2020, Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}. Secondly, the arbitrary mapping between heterographic forms and meaning must also be learned. From the point of view of a modern English speaker, there is no intrinsic reason why \textit{meat} is spelled <ea> and \textit{meet} is spelled <ee>. Nevertheless, these arbitrary spelling distinctions must be learned if they are to be useful, and they presumably place an additional burden on reading and---perhaps even more-so---on writing \parencite{Frith:1979, Shankweiler:1992}.

@@

The second model, \textit{the conservation model} (Fig.~\ref{models_of_heterography}B), explains heterographic homophones as the historical residue of sound change: Two spoken forms merge and become homophonous, but the original spellings are conserved in the orthography. For example, the \textit{meat}--\textit{meet} merger that occurred during the Great Vowel Shift ultimately resulted in Middle English /ɛː/ (spelled <ea>) and /eː/ (spelled <ee>) being pronounced /iː/ in Early Modern English \parencite{Lass:2000}, but the spellings were never changed accordingly, thus giving rise to a set of heterographic homophones that persist in present-day English \parencite[pp.~140--141]{Wells:1982}: \textit{heal}--\textit{heel}, \textit{leak}--\textit{leek}, \textit{meat}--\textit{meet}, \textit{read}--\textit{reed}, \textit{sea}--\textit{see}, \textit{team}--\textit{teem}, \textit{weak}--\textit{week}. The same is true of the \textit{pain}--\textit{pane} merger \parencite[pp.~141--142]{Wells:1982}: \textit{maid}--\textit{made}, \textit{main}--\textit{mane}, \textit{pain}--\textit{pane}, \textit{raise}--\textit{raze}, \textit{sail}--\textit{sale}, \textit{vain}--\textit{vane}. Sound changes involving consonants have also resulted in (or contributed to) pairs of words entering a state of heterography, such as the reduction of /kn/ into /n/ (e.g., \textit{knight}--\textit{night}, \textit{know}--\textit{no}, \textit{knot}--\textit{not}), the loss of /ç/ (e.g., \textit{eight}--\textit{ate}, \textit{right}--\textit{rite}, \textit{sight}--\textit{site}), and the merger of /ʍ/ into /w/ (e.g., \textit{whale}--\textit{wail}, \textit{which}--\textit{witch}, \textit{whine}--\textit{wine}).\footnote{Although we can never be entirely certain how words were pronounced before the advent of sound recording technology, historical linguists have compiled persuasive evidence by a variety of methods. Comparison to modern Dutch and German, for example, offers an insight into how these words might have been pronounced in the past (e.g., \textit{knot} is cognate with German \textit{Knoten} where the /kn/ cluster continues to be fully rendered and \textit{eight} is cognate with Dutch and German \textit{acht}, where the palatal fricative still exists).} A more recent (and perhaps in-progress) example can be found in dialects that have undergone the \textit{father}--\textit{bother} merger, including most varieties of American English \parencite[p.~169]{Labov:2005}, which has resulted in, for example, \textit{balm}--\textit{bomb} (/bɑm/), \textit{lager}--\textit{logger} (/lɑɡər/), and \textit{mach}--\textit{mock} (/mɑk/); these words continue to be spelled with <a> vs.\@ <o> despite being homophonous in such dialects.

@@

All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials\newmaterial{, lasting around 8~min}. Participants were also told explicitly before starting the training session that the stems \newmaterial{looked and sounded like} the objects' shapes, allowing participants to focus on learning the suffixes during the training phase. In passive exposure trials, the alien objects were presented alongside the written and spoken forms in quick succession for 2~s each. In mini-test trials, which were interleaved among the passive exposure trials, the participant was asked to type the appropriate written form for an object and was given feedback on any errors (deleted characters shown in red strikethrough text and additions shown in bold green text). The participant received a 2p bonus for spelling the word correctly but had to submit their response within 20~s. Each of the six object--word pairs in the training set (i.e., the seen items that passed through the bottleneck on transmission) was passively exposed 18 times and mini-tested six times, resulting in a total of 108 passive exposure trials and 36 mini-test trials (the maximum bonus in training was therefore 72p). To check that participants were listening to the spoken forms, they were asked auditorily to click on the alien object at three random points during training; participants who did not follow this instruction were excluded. \newmaterial{The instructions provided to participants are provided in Appendix~A in the supplementary material.}

@@

Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant. Both participants received training on the same orthographic system inherited from the previous generation.\footnote{\newmaterial{Albeit subject to a separate application of the bottleneck on transmission (i.e., the seen items were selected independently for each participant). We felt this was more ecological than giving both participants identical input.}} The communication game, \newmaterial{which shares similarities with \textcite{Kirby:2015},} closely mirrored the overall structure of the test administered to participants in the non-communicative condition described above, with the production and comprehension trials becoming the two sides of a single communicative interaction. On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array). The two participants then switched roles, resulting in the same overall trial structure as the Transmission-only condition (i.e., alternation between production and comprehension trials).

@@

\newmaterial{To analyze how informativeness changes over time and how this compares between the two conditions, we reduced the typological classifications into two broader categories: informative systems (i.e., holistic or expressive) and uninformative systems (i.e., redundant or degenerate). We then fit a Bayesian mixed-effects logistic regression model that predicts whether or not a system is informative as a function of generation with by-chain random slopes and intercepts,\footnote{Model: informative $\sim$ log(generation) + (1 + log(generation) | chain). Bernoulli likelihood with logit link function, weakly informative default priors, six chains of 12000 samples. All statistical models were fit using Bambi 0.13 \parencite{Capretto:2022}. Parameter estimates and diagnostics are provided in Appendix~B of the supplementary material.} which is the standard model structure used to analyze iterated learning experiments \parencite{Winter:2016}. Our Bayesian approach produces posterior estimates of two key parameters: $\alpha$, representing the intercept of the regression model, and $\beta$ representing its slope. To determine if there is a statistical difference between conditions, we compute the difference in slopes, $\Delta(\beta) = \beta_\mathrm{comm} - \beta_\mathrm{trans}$, and check if this posterior difference satisfactorily rejects zero. Here we follow the convention that the 95\% highest density interval (HDI; the narrowest interval that contains 95\% of the posterior probability mass) should not include zero. We emphasize, however, that the posterior is a complete description of the evidence (given the data and model assumptions) and does not strictly need to be reduced to a binary yes/no decision; the further the probability mass is from zero, the stronger the evidence of an effect. The results are shown in Fig.~\ref{results_exp1}B. In both conditions, there is a decrease in informativeness over time (the slopes are negative), but in the Transmission + Communication condition, the slope is shallower, suggesting that informativeness decreases more slowly in the presence of communicative pressure. The difference in slopes, $\Delta(\beta) = 4.53$ (95\%~HDI: 1.48, 7.88), clearly rejects zero, pointing to a meaningful difference between conditions.}

\newmaterial{One issue with the above approach is that collapsing the systems into binary categories (informative vs.\@ uninformative) results in a loss of information about \textit{how informative} the systems are. In addition, there was evidence to suggest that the model was a suboptimal description of the data, since there was also a difference in the $\alpha$ estimates (the intercepts), which should theoretically be the same (i.e., there should be no difference between conditions at Generation~0). We address these limitations with a second measure of informativeness, communicative cost \parencite{KempRegier:2012, Kemp:2018, Regier:2015}, which has previously been used in similar experimental studies \parencite{Carr:2020, Carstensen:2015, Smith:2020} and was proposed in our preregistration as the primary measure of informativeness.} Communicative cost is an information-theoretic measure that expresses how much information will be lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system. If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero~bits---that is, zero information will be lost during each attempt to communicate using that system. Communicative cost will take some larger value if the system contains ambiguity. It is given by $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$, where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$. In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system. \newmaterial{The results are shown in Fig.~\ref{results_exp1}C. We used the same mixed-effects linear regression model described above (except that the likelihood is now Gaussian). In line with the previous analysis, cost increases with generation in both conditions but increases more slowly under communicative pressure. However, the support for a difference between conditions was weaker under this more nuanced measure, $\Delta(\beta) = -0.17$ (95\%~HDI: $-0.36$, $0.02$). Although we were not able to reject zero at the 95\% level, we were able to reject it at the 92\% level (92\%~HDI: $-0.35$, $-0.01$).}

@@

Our first experiment tested whether informative, heterographic orthography could emerge through spelling differentiation and whether it would emerge preferentially under communicative pressure. Although there was evidence to suggest that the orthographic systems remain informative for longer under communicative pressure, both conditions ultimately converged on degenerate, uninformative systems and there was little evidence of systematic differentiation in spelling. Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <ko> for pink, <co> for yellow, and <qo> for blue), but also through less obvious strategies such as the use of length (e.g., <ko> for pink, <koo> for yellow, and <kooo> for blue) or the use of arbitrary silent letters (e.g., <kox> for pink, <kof> for yellow, and <kom> for blue). Such differentiation was not forthcoming, however, even under communicative pressure. Instead, if spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure \parencite[e.g.,][]{Kirby:2015}.

So, what was different about the present experiment compared the large body of prior experimental iterated learning studies? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative about one of the dimensions. Indeed, the point of our experiment is to see whether orthography can resist phonology under sufficient pressure for informativeness. The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color. \newmaterial{In support of this explanation, we conducted an additional experiment during review, which showed that when homophony is removed, the systems tend to resist degeneration in line with prior work. We discuss this experiment in more detail in the Discussion section and in Appendix~C of the supplementary material.}

@@

\newmaterial{The results of both experiments were, however, relatively weak statistically, with the differences in slopes often only just (or not quite) meeting the 95\% criterion. This perhaps reflects a situation of low power, with only ten chains---ten independent sampling units---in each condition. This is a general issue with iterated learning experiments, however, and our use of ten chains follows norms in the field.} It is also important to note, at this point, that the two experiments cannot be compared directly, although we made every effort to keep the two as close as possible. Fundamentally, participants---or more generally, the evolutionary systems---are being asked to do something quite different across the two experiments: create in Experiment~1 and maintain in Experiment~2. The demands of these two tasks are different, and one task or the other may be better suited to our experimental paradigm. Similarly, we cannot draw direct conclusions from the relative strength of the statistical results across the two experiments.

>>>>

\abstract{It is widely acknowledged that opaque orthographies place additional demands on learning, often requiring many years to fully acquire. It is less widely recognized, however, that such opacity may offer certain benefits in the context of reading. For example, heterographic homophones such as <knight> and <night> (words that sound the same but which are spelled differently) impose additional costs in learning but reduce ambiguity in reading. Here, we consider the possibility that---left to evolve freely---writing systems will sometimes choose to forego some simplicity for the sake of informativeness when there is functional pressure to do so. We investigate this hypothesis by simulating the evolution of orthography as it is transmitted from one generation to the next, both with and without a communicative pressure for ambiguity avoidance. In addition, we consider two mechanisms by which informative heterography might be selected for: \textit{differentiation}, in which new spellings are created to differentiate meaning (e.g., <lite> vs.~<light>), and \textit{conservation}, in which heterography arises as a byproduct of sound change (e.g., <meat> vs.~<meet>). Under pressure from learning alone, orthographic systems become transparent, but when combined with communicative pressure, they tend to favor some additional informativeness. Nevertheless, our findings also suggest that, in the long term, simpler, transparent spellings may be preferred in the absence of top-down explicit teaching.}

@@

\noindent
Writing systems, particularly those employing alphabetic scripts, are commonly regarded as providing a visual representation of speech, with letters or chunks of letters corresponding to distinct sounds. However, it is also well understood that writing systems diverge from their spoken counterparts in important ways \parencite{Biber:1988, Bolinger:1946, Coulmas:1991}. The insertion of spacing between words, for example, is almost ubiquitous across alphabetic writing systems, even though no such spacing exists between words in speech \parencite{Parkes:1992, Saenger:1997}. It seems likely that graphic innovations such as these exist because they confer some benefit that is not required in the spoken modality \parencite{Rastle:2019}. In the case of spacing, for example, the separation of words into discrete chunks presumably aids in the targeting and extraction of visuo-linguistic information---constraints that do not exist in the auditory modality. In principle, the same may be true of spelling: Words may be spelled in ways that diverge from the spoken language because such divergence confers some benefit in reading \parencite{Ulicheva:2020}.

One potential case of such functional divergence is heterographic homophony---words that sound alike but which are written differently (e.g., <meat> and <meet> for /miːt/). Heterographic spellings such as these may serve a valuable function in reading. For example, an English speaker faced with a spoken sentence beginning /ðɛr.../ will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr~kat/, a form of the verb \textit{to~be}, as in /ðɛr~ɪz/, or the progressive form of a verb, as in /ðɛr~gəʊɪŋ/. In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate these cases, giving the reader a headstart on processing the upcoming syntactic structure and semantic content. Heterographic homophony is also common below the word level, since many orthographies forego the phonological principle in favor of the morphological principle in the spelling of affixes \parencite{Sandra:2024}. The English suffixes \textit{-er} (denoting the comparative form of an adjective; e.g., \textit{nicer}) and \textit{-or} (denoting the performer of an action; e.g., \textit{actor}) are homophonous in speech (/ər/), but their spellings differentiate these meanings in writing. Of course, English orthography is suboptimal here in~that \textit{-er} may also indicate agentive status (e.g., \textit{builder}); nevertheless, statistical patterns such as these hold across a variety of English affixes \parencite{Berg:2017} and it has been shown that readers are sensitive to and make use of such cues in reading \parencite{Ulicheva:2020}. \newmaterial{Heterography might be especially important given the differing constraints of the written modality, including the lack of other cues to meaning, such as stress, context, and body language, and the inability for reader and writer to engage in immediate feedback and repair. Furthermore, written language has richer vocabulary and more complex syntax than spoken language \parencite{Biber:1988, Korochkina:2024, Nation:2022}, placing different pressures on ambiguity resolution.}

\newmaterial{Heterography is particularly notable in English, but it is also a feature of many other languages and writing systems. French is similar to English in having a large number of heterographic homophones: \textit{cent} (\textit{hundred}), \textit{sang} (\textit{blood}), \textit{sans} (\textit{without}), and \textit{sens} (\textit{feel}), for example, are all pronounced /sɑ̃/ (although their pronunciations will sometimes be distinguished through liaison). In Danish, the words \textit{hver} (\textit{every}), \textit{vejr} (\textit{weather}), \textit{vær} (\textit{be}), and \textit{værd} (\textit{worth}) are all pronounced /vεˀɐ̯/. In Vietnamese, the graphemes <d>, <gi>, and <r> are homophonous, resulting in sets like \textit{dao} (\textit{knife}), \textit{giao} (\textit{delivery}), and \textit{rao} (\textit{advertise}), all pronounced /zāw/. In some cases, features of an orthography designed for other purposes can inadvertently result in homophone disambiguation: Noun capitalization in German contrasts \textit{Wagen} (\textit{car}) and \textit{wagen} (\textit{to dare}), both pronounced /vaːɡṇ/; eclipsis marking in Irish contrasts \textit{bpáistí} (\textit{children}) and \textit{báistí} (\textit{rain}), both pronounced /bɑʃti/; and the morphological principle of Russian orthography contrasts \textit{приступить} (\textit{to start}) and \textit{преступить} (\textit{to transgress}), both pronounced /pristupʲitʲ/. Even in the most transparent of orthographies it is possible to find some instances of heterography: In Italian, the residual~\textsc{h} inherited from Latin contrasts \textit{hanno} (\textit{have}) and \textit{anno} (\textit{year}), both pronounced /anno/, while the grave accent is sometimes used to distinguish common homophonous words, such as \textit{la} (\textit{the}) and \textit{là} (\textit{there}).}

\newmaterial{Perhaps the most elaborate example of how a writing system can deal with homophony head-on is the Chinese orthography. The Chinese spoken languages are rich in homophones, making heterographic spellings---and therefore a logographic writing system---particularly useful \parencite{Frost:2012}. In Mandarin Chinese, the words 糖 (\textit{sugar}), 塘 (\textit{embankment}), 溏 (\textit{pond}), and 搪 (\textit{to block}) are homophonous in speech but heterographic in writing---the phonetic radical on the right (<唐>, /táŋ/) represents the spoken syllable, while the semantic radicals on the left differentiate the meanings \parencite[p.~101]{Coulmas:1991}. In addition, 唐 itself is a surname/dynasty (\textit{Tang}), and another unrelated word 堂 (\textit{hall}) is also pronounced /táŋ/, yielding at~least six ways to write the same sound depending on the meaning. This property allows the written form of Chinese to convey more information about meaning---to be more \textit{informative}---than its spoken counterpart.}

Despite the benefits that heterography may provide in reading, it comes with two main costs. Firstly, by definition, heterography implies that a single sound can be spelled multiple ways. In English, the heterographic spellings <meat> and <meet> imply that /i/ can be spelled <ea> or <ee>. Readers are therefore required to learn alternate spellings for a single sound, \newmaterial{resulting in longer learning periods and more difficult decoding \parencite{Reis:2020, Seymour:2003, Spencer:2003, Taylor:2011, Zhao:2018}}. Secondly, the arbitrary mapping between heterographic forms and meaning must also be learned. From the point of view of a modern English speaker, there is no intrinsic reason why \textit{meat} is spelled <ea> and \textit{meet} is spelled <ee>. Nevertheless, these arbitrary spelling distinctions must be learned if they are to be useful, and they presumably place an additional burden on reading and---perhaps even more-so---on writing \parencite{Frith:1979, Shankweiler:1992}.

@@

The second model, \textit{the conservation model} (Fig.~\ref{models_of_heterography}B), explains heterographic homophones as the historical residue of sound change: Two spoken forms merge and become homophonous, but the original spellings are conserved in the orthography. For example, the \textit{meat}--\textit{meet} merger that occurred during the Great Vowel Shift ultimately resulted in Middle English /ɛː/ (spelled <ea>) and /eː/ (spelled <ee>) being pronounced /iː/ in Early Modern English \parencite{Lass:2000}, but the spellings were never changed accordingly, thus giving rise to a set of heterographic homophones that persist in present-day English \parencite[pp.~140--141]{Wells:1982}: \textit{heal}--\textit{heel}, \textit{leak}--\textit{leek}, \textit{meat}--\textit{meet}, \textit{read}--\textit{reed}, \textit{sea}--\textit{see}, \textit{team}--\textit{teem}, \textit{weak}--\textit{week}. The same is true of the \textit{pain}--\textit{pane} merger \parencite[pp.~141--142]{Wells:1982}: \textit{maid}--\textit{made}, \textit{main}--\textit{mane}, \textit{pain}--\textit{pane}, \textit{raise}--\textit{raze}, \textit{sail}--\textit{sale}, \textit{vain}--\textit{vane}. Sound changes involving consonants have also resulted in (or contributed to) pairs of words entering a state of heterography, such as the reduction of /kn/ into /n/ (e.g., \textit{knight}--\textit{night}, \textit{know}--\textit{no}, \textit{knot}--\textit{not}), the loss of /ç/ (e.g., \textit{eight}--\textit{ate}, \textit{right}--\textit{rite}, \textit{sight}--\textit{site}), and the merger of /ʍ/ into /w/ (e.g., \textit{whale}--\textit{wail}, \textit{which}--\textit{witch}, \textit{whine}--\textit{wine}).\footnote{Although we can never be entirely certain how words were pronounced before the advent of sound recording technology, historical linguists have compiled persuasive evidence by a variety of methods. Comparison to modern German, for example, offers an insight into how these words might have been pronounced in the past (e.g., \textit{knot} is cognate with \textit{Knoten} where the /kn/ cluster continues to be fully rendered and \textit{eight} is cognate with \textit{acht}, where the palatal fricative still exists).} A more recent (and perhaps in-progress) example can be found in dialects that have undergone the \textit{father}--\textit{bother} merger, including most varieties of American English \parencite[p.~169]{Labov:2005}, which has resulted in, for example, \textit{balm}--\textit{bomb} (/bɑm/), \textit{lager}--\textit{logger} (/lɑɡər/), and \textit{mach}--\textit{mock} (/mɑk/); these word pairs continue to be spelled with <a> vs.\@ <o> despite being homophonous in such dialects.

@@

All participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials\newmaterial{, lasting around 8~min}. Participants were also told explicitly before starting the training session that the stems looked and sounded like the objects' shapes, allowing participants to focus on learning the suffixes during the training phase. In passive exposure trials, the alien objects were presented alongside the written and spoken forms in quick succession for 2~s each. In mini-test trials, which were interleaved among the passive exposure trials, the participant was asked to type the appropriate written form for an object and was given feedback on any errors (deleted characters shown in red strikethrough text and additions shown in bold green text). The participant received a 2p bonus for spelling the word correctly but had to submit their response within 20~s. Each of the six object--word pairs in the training set (i.e., the seen items that passed through the bottleneck on transmission) was passively exposed 18 times and mini-tested six times, resulting in a total of 108 passive exposure trials and 36 mini-test trials (the maximum bonus in training was therefore 72p). To check that participants were listening to the spoken forms, they were asked auditorily to click on the alien object at three random points during training; participants who did not follow this instruction were excluded. \newmaterial{The instructions provided to participants are provided in Appendix~A in the supplementary material.}

@@

Participants assigned to the communicative condition completed a live, over-the-internet communication game with another participant. Both participants received training on the same orthographic system inherited from the previous generation.\footnote{\newmaterial{The seen items were selected independently for each participant. We felt this was more ecological than giving both participants identical input.}} The communication game, \newmaterial{which shares similarities with \textcite{Kirby:2015},} closely mirrored the overall structure of the test administered to participants in the non-communicative condition described above, with the production and comprehension trials becoming the two sides of a single communicative interaction. On a given trial, one participant (the director) would complete a production trial under the same input restrictions described above (i.e., produce a form for a target meaning), and the word they used was relayed to the other participant (the matcher), who would then complete a comprehension trial in response to that word (i.e., pick an object from the matcher array). The two participants then switched roles, resulting in the same overall trial structure as the Transmission-only condition (i.e., alternation between production and comprehension trials).

@@

\newmaterial{To analyze how informativeness changes over time and how this compares between the two conditions, we reduced the typological classifications into two broader categories: informative systems (i.e., holistic or expressive) and uninformative systems (i.e., redundant or degenerate). We then fit a Bayesian mixed-effects logistic regression model that predicts whether or not a system is informative as a function of generation with by-chain random slopes and intercepts,\footnote{Model: informative $\sim$ log(generation) + (1 + log(generation) | chain). Bernoulli likelihood with logit link function, weakly informative default priors, six chains of 12000 samples. All statistical models were fit using Bambi 0.13 \parencite{Capretto:2022}. Parameter estimates and diagnostics are provided in Appendix~B of the supplementary material.} which is the standard model structure used to analyze iterated learning experiments \parencite{Winter:2016}. Our Bayesian approach produces posterior estimates of two key parameters: $\alpha$, representing the intercept of the regression model, and $\beta$ representing its slope. To determine if there is a statistical difference between conditions, we compute the difference in slopes, $\Delta(\beta) = \beta_\mathrm{comm} - \beta_\mathrm{trans}$, and check if this posterior difference satisfactorily rejects zero. Here we follow the convention that the 95\% highest density interval (HDI; the narrowest interval that contains 95\% of the posterior probability mass) should not include zero. We emphasize, however, that the posterior is a complete description of the evidence (given the data and model assumptions) and does not strictly need to be reduced to a binary yes/no decision. The results are shown in Fig.~\ref{results_exp1}B. In both conditions, there is a decrease in informativeness over time (the slopes are negative), but in the Transmission + Communication condition, the slope is shallower, suggesting that informativeness decreases more slowly in the presence of communicative pressure. The difference in slopes, $\Delta(\beta) = 4.53$ (95\%~HDI: 1.48, 7.88), clearly rejects zero, pointing to a meaningful difference between conditions.}

\newmaterial{One issue with the above approach is that collapsing the systems into binary categories (informative vs.\@ uninformative) results in a loss of information about \textit{how informative} the systems are. In addition, there was evidence to suggest that the model was a suboptimal description of the data, since there was also a difference in the $\alpha$ estimates (the intercepts), which should theoretically be the same (i.e., there should be no difference between conditions at Generation~0). We address these limitations with a second measure of informativeness, communicative cost \parencite{KempRegier:2012, Kemp:2018, Regier:2015}, which has previously been used in similar experimental studies \parencite{Carr:2020, Carstensen:2015, Smith:2020} and was proposed in our preregistration as the primary measure of informativeness.} Communicative cost is an information-theoretic measure that expresses how much information will be lost, on average, when a speaker/writer attempts to convey a meaning to a listener/reader using some shared signaling system. If the system contains no ambiguities (all meanings are expressed by unique signals), communicative cost will be zero~bits---that is, zero information will be lost during each attempt to communicate using that system. Communicative cost will take some larger value if the system contains ambiguity. It is given by $\sum_{m \in U} \mathrm{Pr}(m) ( -\log \mathrm{Pr}(m|s_m) )$, where $U$ is the universe of meanings that may be expressed, $\mathrm{Pr}(m)$ is the probability that a particular meaning would need to be expressed, and $\mathrm{Pr}(m|s_m)$ is the probability that a reader would infer meaning $m$ given that a writer produced signal $s$ for meaning $m$. In our case, $U$ is the set of nine alien objects, $\mathrm{Pr}(m)$ is set to $1/|U|$ (all objects need to be talked about with equal probability), and $\mathrm{Pr}(m|s_m)$ is given by $1/|M_s|$, where $M_s$ is the set of meanings labeled $s$ according to the system. \newmaterial{The results are shown in Fig.~\ref{results_exp1}C. We used the same mixed-effects linear regression model described above (except that the likelihood is now Gaussian). In line with the previous analysis, cost increases with generation in both conditions but increases more slowly under communicative pressure. However, the support for a difference between conditions was weaker under this more nuanced measure, $\Delta(\beta) = -0.17$ (95\%~HDI: $-0.36$, $0.02$); although we were not able to reject zero at the 95\% level, we were able to reject it at the 92\% level (92\%~HDI: $-0.35$, $-0.01$).}

@@

Our first experiment tested whether informative, heterographic orthography could emerge through spelling differentiation and whether it would emerge preferentially under communicative pressure. Although there was evidence to suggest that the orthographic systems remain informative for longer under communicative pressure, both conditions ultimately converged on degenerate, uninformative systems and there was little evidence of systematic differentiation in spelling. Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <ko> for pink, <co> for yellow, and <qo> for blue), but also through less obvious strategies such as the use of length (e.g., <ko> for pink, <kko> for yellow, and <kkko> for blue) or the use of arbitrary silent letters (e.g., <kox> for pink, <kof> for yellow, and <kom> for blue). Such differentiation was not forthcoming, however, even under communicative pressure. Instead, if spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure \parencite[e.g.,][]{Kirby:2015}.

So, what was different about the present experiment compared to the large body of prior experimental iterated learning studies? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative about one of the dimensions. Indeed, the point of our experiment is to see whether orthography can resist phonology under sufficient pressure for informativeness. The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color. \newmaterial{In support of this explanation, we conducted an additional experiment during review, which showed that when homophony is removed, the systems tend to resist degeneration in line with prior work. We discuss this experiment in more detail in the Discussion section and in Appendix~C of the supplementary material.}

@@

\subsection{Differentiation or conservation?}
